\plaintitle

### 2016-04-22 paper/paper.tex
A bit of work on the paper

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Visualizing NBA teams across multiple seasons}
\def\plainauthor[#1]{#1}
\def\plainauthors[#1]#2#3{#1, #2, #3}
\def\emptyauthor{}
\def\plainkeywords{Authors' choice; of terms; separated; by
  semicolons; include commas, within terms only; required.}
\def\plaingeneralterms{Documentation, Standardization}

@@

This paper describes the visualization of NBA data made by the ``The Tufters''
team for the course ``Information Visualization''. In section~\ref{sec:goal}
we describe the goal of the visualization and the target audience. In
section~\ref{sec:data} we describe the data used, including its origins,
advantages and limitations. In section~\ref{sec:literature} we give an overview
of related literature and web resources, including related visualizations.
Related visualizations include both visualizations of NBA or other sports data,
as well as visualizations tackling a problem we encountered during the
development of our visualization. In section~\ref{sec:visualization} we describe
the visualization itself. We give an overview of the different stages of
development of the visualization, as well as the major design decisions made
during the development process. Section~\ref{sec:discussion} discusses potential
improvements of the final visualization and lessons learned from the project. We
conclude in section~\ref{sec:conclusion} by giving a short overview.

\section{Goal and Target Audience}\ref{sec:goal}
The visualization's goal is allowing exploration of NBA data by lay persons.
More specifically, the visualization does not try to offer premade explanations
for phenomenons visible in the NBA data. By providing easy and intuitive access
to the data, users can draw their own conclusions. It thus enables users of all 
sorts to explore the data for patterns.
Secondly, the visualization's target audience are lay persons. This means it 
satifies needs of people regardless if they are professionally active in the 
field of basketball or not. Specifically, fans of NBA are the core of our target
audience. This means the visualization assumes most common basketball terms are
known to the audience and as such does not provide additional information about 
them.
The visualization should enable people to look for changes in team statistics 
and evaluate the impact on the team. The visualization gives also a short insight
in what caused the change in statistics. Eg. a user could notice a spike/drop in 
the SRS score () of a team. The user will then be able to see how the spike was 
caused (team performance, players that joined the team, players that left the 
team). Next to that, the user will can also explore the impact of the change on
the team. She/he will be able to scroll through time and notice how the results of
the team change over time since the spike/drop.

\section{Data}\ref{sec:data}
The data visualized is a subset of the data available on basketball statistics
site basketball-reference\cite{basketball-reference}. A wide range of data is
available on this site. In our visualization we only use data from 1984 onwards.
The data we use includes league standings and playoff rankings for each team,
team overall statistics, the team's roster and individual player statistics 
for each year, including the PER~(Player Efficiency Rating).

The data was gathered by scraping the basketball-reference site, mostly using
the provided download capabilities. Most of the data was downloaded in csv
format, while some tables had to be manually scraped. The data was then combined
in a preprocessing step. In this step, each team's playoff rankings were
calculated based on the matches played during the playoffs, and the rest of the
data was combined into json format. The final preprocessing step combines all
data into one json file.

\section{Related Work}\ref{sec:literature}

\section{Visualization}\ref{sec:visualization}
The visualization consists of three parts:
- The bubble view: a compact view on the play-offs per season
- The statistics (zoom) view: a more detailed view on a statistic of a selected 
team
- The team view: a detailed view on how good or bad a team scores on a specific 
field position
All three of them are discussed in more detail with the rationale behind them.
The webpage is built in such a way that it gives a hint to the user at the bottom
or the top of a page when another view is available. By showing a different 
background color and an arrow, we invite the visitor to explore other views. The 
user is also at liberty to scroll through these views. However, a fixed scroll is
implemeted so users will only end in complete views when moving around. Note that 
the user also will be guided through all 3 views when performing actions described 
in the section below.

@@

\subsection{The statistics view}
The statistics view gives a user a clear overview of how a statistic has been 
influenced compared to the year before. The view consists of 3 parts:
- Context section: the top part should inform the user of his/her context.
- An arrow section illustrating why a statistic changed
- Small multiples of team statistics and how they evolve over time
\subsubsection{Context section}
The context section should inform the user at all times of the current selections 
that have been made. The statistics view will always be based on a specific team
and season that have been selected or that are initated. This should be clear based
on the textual information in the top left. Next to that, in the top right a small 
bubble view is shown. It indicates the team selected and how well it performed in 
the play-offs during that season. When changing teams and/or season, this view gets 
updated as well as the textual information.

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a choosen statistic compared with the season
the year before. This should give a visitor more insight into why a statistic changed
over time. The change is represented by three arrows:
- arrow on the left pointing towards the middle: indicating the influence of players
who joined the team
- arrow in the middle: indicating how the team inherent changed
- arrow on the right pointing towards the middle: indicating the influence of players
who left the team
<figure of the statistics view>
The shirts next to the arrows are the shirts of players that had the biggest 
contribution to that arrow. Note that these shirts don't change in size. The box on 
the right enables the user to change between 
statistics. Next to that it informs the user of which statistic has been choosen. When 
the user hits the circle in the middle representing the teams SRS scrore, he is guided 
to the team view. This view will be explained in more detail in the next section.

@@

\section{Lessons learned}\ref{sec:discussion}
We've learned a couple of lessons during the creation of this visualization:
- Find a use case as soon as possible to built your visualization around
- Each separate component should be a sufficient informative visualization on itself
- Creating a custom visualization costs time and opportunities
In what follows we will discuss our experience in more detail.

@@

\section{Conclusion}\ref{sec:conclusion}

@@

>>>>

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Why did they win?\\Visualizing NBA teams across multiple seasons}
\def\plainauthor[#1]{#1}
\def\plainauthors[#1]#2#3{#1, #2, #3}
\def\emptyauthor{}
\def\plainkeywords{Information Visualization; CHI; NBA}
\def\plaingeneralterms{}

@@

This paper describes the visualization of NBA data made by the ``The Tufters''
team for the course ``Information Visualization''. In section~\ref{sec:goal} we
describe the goal of the visualization and the target audience. In
section~\ref{sec:data} we describe the data used, including its origins,
advantages and limitations. In section~\ref{sec:literature} we give an overview
of related literature and web resources, including related visualizations.  This
includes both visualizations of NBA or other sports data, as well as
visualizations tackling an isue encountered during the development of our
visualization. In section~\ref{sec:visualization} we describe the visualization
itself. We give an overview of the different stages of development of the
visualization, as well as the major design decisions.
Section~\ref{sec:discussion} discusses potential improvements of the final
visualization and lessons learned from the project. We conclude in
section~\ref{sec:conclusion}. Appendix~\ref{sec:terminology} contains
definitions for common basketball terms used in this paper.

\section{Goal and Target Audience}\label{sec:goal} 
The visualization's goal can be best summarized by the following sentence: ``Why
did they win?'' The visualization focuses on exposing relations between team
performance and their player roster over several years. It allows users to find
explanations for major improvements and declines in team performance.  The
visualization allows exploration of NBA data by lay persons.  More specifically,
the visualization does not try to offer premade explanations for phenomenons
visible in the NBA data. By providing easy and intuitive access to the data, the
visualization allows users to draw their own conclusions. The visualization's
target audience are lay persons. Specifically, fans of NBA are the core of our
target audience. This means the visualization assumes most common basketball
terms are known to the audience and does not provide basic terminology
explanations.

\section{Data}\label{sec:data}
The data visualized is a subset of the data available on basketball statistics
site basketball-reference\cite{basketball-reference}. A wide range of data is
available on this site. This includes common basketball performance statistics,
such as field goals, percentage of shots scored, shooting distance, minutes
played per season, number of personal fouls made, salaries and much more. These
statistics are avalable for each player individually for each season. Aggregated
data for entire teams and for entire seasons is also available. Additionally,
idividual game statistics are also available, including playoff games.  In our
visualization we only use data from 1984 onwards. This is done because of rule
changes in the NBA which changed the number of teams competing and the
competition's structure. To simplify implementation of our visualization, only
data after the last major rule change was used.  The data we use includes league
standings and playoff rankings for each team, team overall statistics, the
team's roster and individual player statistics for each year, including the
PER~(Player Efficiency Rating)\cite{per}.

The data was gathered by scraping the basketball-reference site using the
provided download capabilities. Most of the data was downloaded in csv format,
while some tables had to be manually scraped. This process was automated using
Python scripts. The data was then combined in a preprocessing step. In this
step, each team's playoff rankings were calculated based on the matches played
during the playoffs, and the rest of the data was combined into json format. The
final preprocessing step combined all data into one json file.

\section{Related Work}\label{sec:literature}

\section{Visualization}\label{sec:visualization}
The visualization consists of three parts:
\paragraph{The bubble view} a compact view on the play-offs per season
\paragraph{The statistics (zoom) view} a more detailed view on a statistic of a selected 
team
\paragraph{The team view} a detailed view on how good or bad a team scores on a specific 
field position
All three are discussed in more detail below. The three views are available to
the user on a single webpage. The user start on the bubble view and is able to
navigate to the statistics and team views by scrolling down or clicking on
teams.
%The webpage is built in such a way that it gives a hint to the user at the bottom
%or the top of a page when another view is available. By showing a different 
%background color and an arrow, we invite the visitor to explore other views. The 
%user is also at liberty to scroll through these views. However, a fixed scroll is
%implemeted so users will only end in complete views when moving around. Note that 
%the user also will be guided through all 3 views when performing actions described 
%in the section below.

@@

\subsection{The statistics view}
The statistics view gives a user a clear overview of how a statistic has been 
influenced compared to the year before. The view consists of 3 parts:
\paragraph{Context section} the top part should inform the user of his/her context.
\paragraph{Arrow section} This section illustrates why a statistic changed
\paragraph{Small multiples section} Small multiples of team statistics and how they evolve over time
\subsubsection{Context section}
The context section should inform the user at all times of the current selections 
that have been made. The statistics view will always be based on a specific team
and season that have been selected or that are initated. This should be clear based
on the textual information in the top left. Next to that, in the top right a small 
bubble view is shown. It indicates the team selected and how well it performed in 
the play-offs during that season. When changing teams and/or season, this view gets 
updated as well as the textual information.

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a choosen statistic compared with the season
the year before. This should give a visitor more insight into why a statistic changed
over time. The change is represented by three arrows:
\begin{itemize}
    \item arrow on the left pointing towards the middle: indicating the influence of players
    who joined the team
    \item arrow in the middle: indicating how the team inherent changed
    \item arrow on the right pointing towards the middle: indicating the influence of players
who left the team
\end{itemize}
<figure of the statistics view>
The shirts next to the arrows are the shirts of players that had the biggest 
contribution to that arrow. Note that these shirts don't change in size. The box on 
the right enables the user to change between 
statistics. Next to that it informs the user of which statistic has been choosen. When 
the user hits the circle in the middle representing the teams SRS scrore, he is guided 
to the team view. This view will be explained in more detail in the next section.

@@

\section{Lessons learned}\label{sec:discussion}
We've learned a couple of lessons during the creation of this visualization:
- Find a use case as soon as possible to built your visualization around
- Each separate component should be a sufficient informative visualization on itself
- Creating a custom visualization costs time and opportunities
In what follows we will discuss our experience in more detail.

@@

\section{Conclusion}\label{sec:conclusion}

@@

\begin{appendices}

    \section{Terminology}\label{sec:terminology}

    \paragraph{PER} Player Efficiency Rating - An all-in-one basketball rating,
    boiling down all of a player's contributions into one number\cite{per}.
\end{appendices}
